---
title: |
  bpnreg: A package to analyze Bayesian projected normal circular regression
  models
tags:
  - R
  - Bayesian statistics
  - Circular data
  - Regression
  - Mixed-effects models
authors:
  - name: Jolien Cremers
    orcid: 0000-0002-0037-1747
    affiliation: 1
affiliations:
 - name: Section of Biostatistics, Department of Public Health, University of Copenhagen
   index: 1
date: "24/02/2020"
bibliography: paper.bib
output:
  md_document:
    preserve_yaml: true
    variant: markdown
    pandoc_args: "--atx-headers"
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

The analysis of circular data, measured in degrees or radians, is inherently
different from standard linear data due to its periodic nature ($0^\circ =
360^\circ$). Special methods are needed, even for computing basic summary
statistics [@mardia2009directional]. Although circular data is not very common
they occur in a diverse range of research areas, astronomy, medicine, genetics,
neurology, environmetrics, psychology and geology among others. Software
implementations for the analysis of regression type models for circular data are
however scarce.

To date `bpnreg` is the only software package that implements methods to analyse
both regression and mixed-effects models for circular outcomes.


# Statement of Need

The purpose of `bpnreg` is to provide methods for fitting circular regression
and mixed-effects models to `R`-users. Its two main functions allow for fitting
Bayesian multiple and mixed-effect regression models for circular data based on
the projected normal distribution (see @Nunez-Antonio2011-fm and
@Nunez-Antonio2014-bd for a description of the models). Both continuous and
categorical predictors can be included. Sampling from the posterior is performed
via an MCMC algorithm implemented in `c++` that allows for fast computation (see
@Cremers2018-ta  and @Cremers2021-mm  for a description and assessment of the
MCMC samplers). Posterior descriptives of all parameters, model fit statistics
and Bayes factors for hypothesis tests for inequality constrained hypotheses are
provided.

There are several R-packages that provide methods for circular data analyses,
but only a couple contain functionality for circular regression models (see
@pewsey2020recent for an overview). The package `circular` [@Agostinelli2017] is
a general purpose package that also contains functionality for fitting
frequentist regression models for circular outcomes based on the von Mises
distribution. The package `circglmbayes` [@Mulder2017] provides a Bayesian
regression model for circular outcomes based on the von Mises distribution. To
date however, `bpnreg` is the only R-package providing Bayesian multiple
regression as well as mixed-effects models for circular outcomes based on the
projected normal distribution.

From its first release several published articles [@Tyson-Carr2020-mu;
@Cremers2021-mm; @Cote2020-xg; @Rafferty2020-pb; @Ojeda_undated-by;
@Olson2020-al; @Klugkist2018-ag; @Spinks2019-ya] have used `bpnreg` for circular
data analysis.

# Use

To use `bpnreg` the user first needs install the package from `CRAN` and load the
package as follows:

```{r message=FALSE, eval = FALSE}
install.packages("bpnreg")
library(bpnreg)
```

```{r message=FALSE, echo = FALSE}
library(bpnreg)
```

The two main functions are `bpnr` for fitting Bayesian projected normal multiple
regression models for circular outcomes and `bpnme` for fitting Bayesian
projected normal mixed-effects models for circular outcomes. For detailed
examples on how to fit these two types of models using `bpnreg` and circular
data analysis in general we refer to @Cremers2018-ys. Answers to frequently
asked questions regarding the use of `bpnreg` can be found in the `FAQ` vignette
available on `CRAN`.


# Acknowledgements
JC is supported for this work by a research grant from the Novo Nordisk Foundation (“Harnessing The Power of Big Data to Address the Societal Challenge of Aging.” NNF17OC0027812)


# References
