---
title: |
  bpnreg: An R-package to analyze Bayesian projected normal circular regression
  and mixed-effects models
tags:
  - R
  - Bayesian statistics
  - Circular data
  - Regression
  - Mixed-effects models
authors:
  - name: Jolien Cremers
    orcid: 0000-0002-0037-1747
    affiliation: 1
affiliations:
 - name: Section of Biostatistics, Department of Public Health, University of Copenhagen
   index: 1
date: "24/02/2020"
bibliography: paper.bib
output:
  md_document:
    preserve_yaml: true
    variant: markdown
    pandoc_args: "--atx-headers"
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

Circular data is measured in degrees or radians and could for example consist of
compass directions or clock times. Its analysis is inherently different from
standard linear data due to its periodic nature ($0^\circ = 360^\circ$ and 24:00
= 00:00). Special methods are needed, even for computing basic summary
statistics [@mardia2009directional]. Although circular data are not very common
they occur in a diverse range of research areas, astronomy, medicine, genetics,
neurology, environmetrics, psychology and geology among others. Software
implementations for the analysis of regression and mixed-effects models for
circular data are however scarce. To date `bpnreg` is the only software package
that implements methods to analyze both regression and mixed-effects models for
circular outcomes.


# Statement of Need

The purpose of `bpnreg` is to provide methods for fitting circular regression
and mixed-effects models to `R`-users. Its two main functions allow for fitting
Bayesian multiple and mixed-effects regression models for circular data based on
the projected normal distribution (see @Nunez-Antonio2011-fm and
@Nunez-Antonio2014-bd for a description of the models). Both continuous and
categorical predictors can be included. Sampling from the posterior is performed
via an MCMC algorithm implemented in `c++` that allows for fast computation (see
@Cremers2018-ta  and @Cremers2021-mm  for a description and assessment of the
MCMC samplers). 

There are several R-packages that provide methods for circular data analyses,
but only a couple contain functionality for circular regression models (see
@pewsey2020recent for an overview). The package `circular` [@Agostinelli2017] is
a general purpose package that also contains functionality for fitting
frequentist regression models for circular outcomes based on the von Mises
distribution. The package `circglmbayes` [@Mulder2017] provides a Bayesian
regression model for circular outcomes based on the von Mises distribution. To
date however, `bpnreg` is the only R-package providing Bayesian regression as
well as mixed-effects models for circular outcomes based on the projected
normal distribution.

From its first release `bpnreg` has been used for circular data analysis in
several publications [@Tyson-Carr2020-mu;
@Cremers2021-mm; @Cote2020-xg; @Rafferty2020-pb; @Ojeda_undated-by;
@Olson2020-al; @Klugkist2018-ag; @Spinks2019-ya].


# Example

@Brunye2015-kb studies cognitive maps, defined as map-like memories of familiar
environments that are used when we try to navigate in a certain environment.
Their hypothesis of interest is that similar to physical maps, cognitive maps
also have a north-up orientation. The north-up hypothesis was tested by asking
university students to point North at several campus locations. The variable of
interest, the pointing direction of the students, was measured on a compass and
is thus a circular variable. See @Brunye2015-kb and
@Cremers2018-ta for a more detailed description of the data.

To test whether this pointing direction is affected by a set of covariates a circular
regression model can be used. To use `bpnreg` for fitting such a circular
regression model, the user first needs install the package from `CRAN` and load
the package as follows:

```{r message=FALSE, eval = FALSE}
install.packages("bpnreg")
library(bpnreg)
```

```{r message=FALSE, echo = FALSE, warning = FALSE}
library(bpnreg)
```

```{r, echo = FALSE, results = FALSE}
data <- read.table("DatatoSamplerPointingNorth.txt", header = TRUE, sep = "\t", dec = ".")
data$sex <- factor(data$sex, labels = c("male", "female"))
```

Subsequently, a circular regression model with the covariances sex, age,
experience living on campus and a sense of direction score (sbsod) can be fit to
the data using the `bpnr` function as follows:

```{r, warning = FALSE, message = FALSE, results = FALSE}
fit <- bpnr(theta ~ sex + age + experience + sbsod,
            data = data,
            its = 10000, burn = 100,
            seed = 101)
```

The first argument contain the regression equation (theta is the circular
outcome), in the second argument the data is defined, the following arguments
are used to set the amount of iterations (`its`) and burn-in period (`burn`) for
the MCMC algorithm and finally a seed is set to make the code reproducible. The
results from the circular regression are stored in a new object `fit`.

The package `bpnreg` contains several functions for evaluating the results. The
function `coef_circ` can be used to obtain summary statistics (mean, mode,
standard deviation and 95\% highest posterior density interval) of the posterior
samples of circular coefficients for categorical and continuous covariates
(`type` argument) and can be displayed in both degrees and radians (`units`
argument). We obtain posterior summaries for our regression model as follows:

```{r}
coef_circ(fit, type = "categorical", units = "degrees")
```

```{r, eval = FALSE}
coef_circ(fit, type = "continuous", units = "degrees")
```

```{r, echo = FALSE}
round(coef_circ(fit, type = "continuous", units = "degrees"), 2)
```

As can be seen, posterior summaries for the circular means for males (intercept)
and females as well as the difference in means is displayed. For the three
continuous covariates in the model three types of circular regression
coefficients, `bc`, `AS` and `SAM`, are displayed. For more information on the
interpretation of the results we refer to @Cremers2018-ys.

Functionality for obtaining fit statistics for model comparison is implemented
in the `fit` function:

```{r}
fit(fit)
```

Finally, `traceplot` is a useful function  to evaluate sampler convergence. E.g.
traceplots for the circular regression coefficient `bc` for all continuous
variables in the model are obtained as follows:

```{r}
traceplot(fit, parameter = "b.c")
```
If a traceplot for a specific variable is needed the `variable` argument can be
used. For example:

```{r}
traceplot(fit, parameter = "b.c", variable = "age")
```

For more detailed examples on how to fit circular regression and mixed-effects
models using `bpnreg` and circular data analysis in general we refer to
@Cremers2018-ys. Answers to frequently asked questions regarding the use of
`bpnreg` can be found in the `FAQ` vignette available on `CRAN`.


# Acknowledgements
JC is supported for this work by a research grant from the Novo Nordisk
Foundation (“Harnessing The Power of Big Data to Address the Societal Challenge
of Aging.” NNF17OC0027812).


# References
