---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```
# bpnreg
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/bpnreg)](https://cran.r-project.org/package=bpnreg) 
[![Build Status](https://travis-ci.org/joliencremers/bpnreg.svg?branch=master)](https://travis-ci.org/joliencremers/bpnreg)

The goal of bpnreg is to fit Bayesian projected normal regression models for circular data.

## Installation


The R-package bpnreg can be installed from CRAN as follows:

```{r gh-installation cran, eval = FALSE}
install.packages("bpnreg")
```

You can install  a beta-version of bpnreg from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("joliencremers/bpnreg")
```

## Citation

To cite the package ‘bpnreg’ in publications use:

  Jolien Cremers (2020). bpnreg: Bayesian Projected Normal Regression Models for
  Circular Data. R package version 2.0.1. https://CRAN.R-project.org/package=bpnreg

## Example

This is a basic example which shows you how to run a Bayesian projected normal regression model:

```{r example}
library(bpnreg)
bpnr(Phaserad ~ Cond + AvAmp, Motor)
```
